---
title: "Predictions: Weight Lifting Exercises Dataset"
author: "Roberto Martinez de Morentin"
date: "26/09/2019"
output:
  pdf_document: default
  html_document:
    keep_md: yes
---

## Overview

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit, it is now possible to collect a large amount of data about personal activity relatively inexpensively. The aim of this project is to predict the manner in which participants perform a barbell lift. The data comes from http://groupware.les.inf.puc-rio.br/har wherein 6 participants were asked to perform the same set of exercises correctly and incorrectly with accelerometers placed on the belt, forearm, arm, and dumbell.  

The following steps are followed:

1. Data Preprocessing
2. Exploratory Analysis
3. Prediction Model Selection
4. Predicting Test Set Output

## Data Preprocessing 

Load the training and testing set from the online sources and then split the training set further into training and test sets. 

```{r DataLoading, message = FALSE}

library(caret)
setwd("~/GitHub/PracticalMachineLearning/")
train.url <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
test.url <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

training <- read.csv(url(train.url))
testing <- read.csv(url(test.url))

label <- createDataPartition(training$classe, p = 0.7, list = FALSE)
train <- training[label, ]
test <- training[-label, ]

```

Since from 160 variables present in the dataset, some variables have nearly zero variance while others contain a lot of NA terms which need to be excluded from the dataset. Let’s remove those.


```{r DataCleaning}

nzv <- nearZeroVar(train)
train <- train[ ,-nzv]
test <- test[ ,-nzv]

label <- apply(train, 2, function(x) mean(is.na(x))) > 0.95
train <- train[, -which(label, label == FALSE)]
test <- test[, -which(label, label == FALSE)]

train <- train[ , -(1:5)]
test <- test[ , -(1:5)]

```

We have reduced 160 variables to 54.

## Exploratory Analysis

After cleaneing the dataset off absolutely useless variables, we look at the correlation of these variables. 

```{r CorrelationPlot, fig.width=12, fig.height=8}

library(corrplot)
corrMat <- cor(train[,-54])
corrplot(corrMat, method = "circle", type = "lower", tl.cex = 0.8, tl.col = rgb(0,0,0))

```

Darker gradient correspond to having high correlation. A PCA could be run to further reduce the correlated variables but it not neccessary due to the few number of correlations.

## Prediction Model Selection

We are going to use 3 methods to model the training set and thereby choose the one having the best accuracy to predict the outcome variable in the testing set. The methods will be Decision Tree, Random Forest and Generalized Boosted Model.

A confusion matrix plotted at the end of each model will help us to visualize each analysis.

### Decision Tree

```{r DecisionTree, message = FALSE, warning = FALSE, fig.width=18, fig.height=10}

library(rpart)
library(rpart.plot)
library(rattle)
set.seed(9306)
modelDT <- rpart(classe ~ ., data = train, method = "class")
rpart.plot(modelDT)

predictDT <- predict(modelDT, test, type = "class")
confMatDT <- confusionMatrix(predictDT, test$classe)
confMatDT

```

### Random Forest

```{r RandomForest, message = FALSE}

library(caret)
set.seed(9306)
control <- trainControl(method = "cv", number = 3, verboseIter=FALSE)
modelRF <- train(classe ~ ., data = train, method = "rf", trControl = control)
modelRF$finalModel

predictRF <- predict(modelRF, test)
confMatRF <- confusionMatrix(predictRF, test$classe)
confMatRF

```

### Generalized Boosted Model

```{r GBM, message = FALSE}

library(caret)
set.seed(9306)
control <- trainControl(method = "repeatedcv", number = 5, repeats = 1, verboseIter = FALSE)
modelGBM <- train(classe ~ ., data = train, trControl = control, method = "gbm", verbose = FALSE)
modelGBM$finalModel

predictGBM <- predict(modelGBM, test)
confMatGBM <- confusionMatrix(predictGBM, test$classe)
confMatGBM

```

As Random Forest offers the maximum accuracy of 99.74%, we will go with Random Forest Model to predict our test data class variable.

## Predicting Test Set Output

```{r TestSetPrediction, messages = FALSE}

predictRF <- predict(modelRF, testing)
predictRF

```
